Identifying Software Performance Changes Across Variants and Versions

### 2020-08-14 paper.tex
changes

		In this work, we combine two perspectives -- variability and time -- {\color{red} into a novel perspective}. We propose an approach to identify configuration-dependent performance changes retrospectively across software variants and versions. In a nutshell, we iteratively sample pairs of configurations and versions and measure the respective performance that help us update a model of likelihoods for performance changes. Pursuing a search strategy with the goal of measuring selectively and incrementally further pairs, we increase the accuracy of identified change points related to configuration options and interactions.

		We have conducted a number of experiments both on controlled synthetic datasets as well as in real-world scenarios with different software systems. Our evaluation demonstrates that we can pinpoint performance shifts to configuration options and interactions as well as commits introducing change points with high accuracy and at scale. Our experiments on three real-world systems {\color{red}explore} the effectiveness and practicality of our approach.

@@

	{\color{red}Our goal is to enable the \emph{retrospective} detection of changes in the performance of configurable software systems and pinpoint them to a specific option or interaction. For example, a patch of a certain feature is likely to affect only configurations where such feature is selected. Many software systems are configurable, but have no performance regression testing routine set in place. To uncover performance deficiencies that emerged from revisions in the past development history, it is infeasible to test each and every commit and configuration.} In essence, we face a huge combinatorial problem for performance-change detection: code changes across software versions and behavioral changes across software variants.
	Our problem arises along two dimensions: time (versions) and configuration space (variants).
	\emph{First}, we aim at finding performance changes in software performance from one version to another.
	The detection of such changes is referred to as \emph{change point detection} and is a recurring theme in temporal analysis.
	The main limiting factor is that exhaustive measurements across the configuration space are neither available nor feasible~\cite{white_selecting_2009}.
	Change point detection techniques with both exhaustive measurement~\cite{cityIdentifying2014,daly_industry_2020} as well as limited data availability~\cite{sandoval_alcocer_learning_2016,alcocer_prioritizing_2020,huang_performance_2014,muhlbauer_accurate_2019} have been successfully applied to identify performance changes.
	\emph{Second}, we aim at associating changes in performance (from one version to another) with particular configuration options or interactions among them.
	There is substantial work  regarding a related problem, which is estimating the influence of individual options on performance~\cite{siegmundPredictingPerformanceAutomated2012,siegmundPerformanceinfluenceModelsHighly2015,sarkarCostEfficientSamplingPerformance,haDeepPerf2019}. Instead of estimating the influence of options and interactions on performance, we want to know: Which option or interaction is \emph{responsible} for a particular change point?

@@

	Our main idea is as follows: We address the configuration complexity of this problem by selecting representative sample sets of configurations.
	Next, we uniformly sample a constant number of commits for each configuration and conduct respective performance measurements.
	Based on these measurement data, we assemble a prediction model.
	For each configuration, we estimate the likelihood of each commit being a change point, i.e. that performance of some configurations changes abruptly compared to the previous commit.
	Next, we leverage similarities in the performance histories of configurations that share common options.
	Since we sample the commits for each configuration independently, for each change point, we obtain many estimations using measurements of different commits.
	Overall, this allows us to obtain more accurate estimations of performance-changing commits with tractable effort.
	Based on a mapping from configurations to predicted change point probabilities for each configuration and commit, we derive the options responsible for each particular change point: configuration-dependent change points.
	In summary, we offer the following contributions:
	\begin{itemize}
		\item A {\color{red}novel perspective and} technique to effectively identify shifts in performance of configurable software systems. We are able to pinpoint causative commits and affected configuration options with high accuracy and tractable effort.
		\item A feasibility demonstration of our approach by implementing an adaptive learning strategy to obtain accurate estimations with acceptable measurement cost.
		\item An evaluation using both synthetic and real-world performance data from three configurable software systems. Synthetic data lets us assess our model and approach conceptually and at scale, whereas we are able to assess practicality with real-world data.
		\item A companion website\footnote{\url{https://github.com/AI-4-SE/Changepoints-Across-Variants-And-Versions} or archived at {\color{red}\url{SOFTWAR-HERITAGE}}} providing supplementary material including a reference implementation of our approach, performance measurement data, and additional visualizations.
	\end{itemize}

@@

@@

 	\subsubsection{Acquiring Configurations}
 	The exploration of new configurations is similar to the initial sample selection strategy (distance-based sampling) described in Section~\ref{sec:initialization}. We select a constant number of configurations (default value: 75) for exploration. The exploitation of new configurations requires additional constraints. We define the following constraints for exploiting the candidate solution's associations. For each commit in the candidate solution, we perform exploitation in two directions. For the first direction, at least 80\,\% of  configuration options associated with a commit are required to be selected, whereas, for the second direction, at least 80\,\% configuration options not associated with a commit are required to be selected. This way, we limit variation among configuration options. The first direction aims at enabling additional configuration options to search for previously unseen associations (under-approximation). The second direction aims at the opposite: We limit variation to those configuration options already associated with a commit and therefore search for configuration options that are falsely associated (over-approximation). We use the limit of 80\,\% to allow a broader scope of valid configurations to be sampled by a constraint solver. For each new configuration, we assess performance for randomly selected commits.

>>>>

		In this work, we combine two perspectives -- variability and time --  into a novel perspective. We propose an approach to identify configuration-dependent performance changes retrospectively across software variants and versions. In a nutshell, we iteratively sample pairs of configurations and versions and measure the respective performance that help us update a model of likelihoods for performance changes. Pursuing a search strategy with the goal of measuring selectively and incrementally further pairs, we increase the accuracy of identified change points related to configuration options and interactions.

		We have conducted a number of experiments both on controlled synthetic datasets as well as in real-world scenarios with different software systems. Our evaluation demonstrates that we can pinpoint performance shifts to configuration options and interactions as well as commits introducing change points with high accuracy and at scale. Our experiments on three real-world systems explore the effectiveness and practicality of our approach.

@@

	Our goal is to enable the \emph{retrospective} detection of changes in the performance of configurable software systems and pinpoint them to a specific option or interaction. For example, a patch of a certain feature is likely to affect only configurations where such feature is selected. Many software systems are configurable, but have no performance regression testing routine set in place. To uncover performance deficiencies that emerged from revisions in the past development history, it is infeasible to test each and every commit and configuration. In essence, we face a huge combinatorial problem for performance-change detection: code changes across software versions and behavioral changes across software variants.
	Our problem arises along two dimensions: time (versions) and configuration space (variants).
	\emph{First}, we aim at finding performance changes in software performance from one version to another.
	The detection of such changes is referred to as \emph{change point detection} and is a recurring theme in temporal analysis.
	The main limiting factor is that exhaustive measurements across the configuration space are neither available nor feasible~\cite{white_selecting_2009}.
	Change point detection techniques with both exhaustive measurement~\cite{cityIdentifying2014,daly_industry_2020} as well as limited data availability~\cite{sandoval_alcocer_learning_2016,alcocer_prioritizing_2020,huang_performance_2014,muhlbauer_accurate_2019} have been successfully applied to identify performance changes.
	\emph{Second}, we aim at associating changes in performance (from one version to another) with particular configuration options or interactions among them.
	There is substantial work  regarding a related problem, which is estimating the influence of individual options on performance~\cite{siegmundPredictingPerformanceAutomated2012,siegmundPerformanceinfluenceModelsHighly2015,sarkarCostEfficientSamplingPerformance,haDeepPerf2019}. Instead of estimating the influence of options and interactions on performance, we want to know: Which option or interaction is \emph{responsible} for a particular change point?

@@

	Our main idea is as follows: We address the configuration complexity of this problem by selecting representative sample sets of configurations.
	Next, we uniformly sample a constant number of commits for each configuration and conduct respective performance measurements.
	Based on these measurement data, we assemble a prediction model.
	For each configuration, we estimate the likelihood of each commit being a change point, i.e. that performance of some configurations changes abruptly compared to the previous commit.
	Next, we leverage similarities in the performance histories of configurations that share common options.
	Since we sample the commits for each configuration independently, for each change point, we obtain many estimations using measurements of different commits.
	Overall, this allows us to obtain more accurate estimations of performance-changing commits with tractable effort.
	Based on a mapping from configurations to predicted change point probabilities for each configuration and commit, we derive the options responsible for each particular change point: configuration-dependent change points.
	In summary, we offer the following contributions:
	\begin{itemize}
		\item A novel technique to effectively identify shifts in performance of configurable software systems. We are able to pinpoint causative commits and affected configuration options with high accuracy and tractable effort.
		\item A feasibility demonstration of our approach by implementing an adaptive learning strategy to obtain accurate estimations with acceptable measurement cost.
		\item An evaluation using both synthetic and real-world performance data from three configurable software systems. Synthetic data lets us assess our model and approach conceptually and at scale, whereas we are able to assess practicality with real-world data.
		\item A companion website\footnote{\url{https://github.com/AI-4-SE/Changepoints-Across-Variants-And-Versions} or archived at {\color{red}\url{SOFTWAR-HERITAGE}}} providing supplementary material including a reference implementation of our approach, performance measurement data, and additional visualizations.
	\end{itemize}

@@

	\begin{figure}
		\includegraphics[width=0.5\textwidth]{images/sampling_plan}
		\label{fig:sampling plan}
		\caption{Active sampling strategies for a DBMS with 4 configurations}
	\end{figure}

@@

	\subsubsection{Acquiring Configurations}
	The exploration of new configurations is similar to the initial sample selection strategy (distance-based sampling) described in Section~\ref{sec:initialization}. We select a constant number of configurations (default value: 75) for exploration. The exploitation of new configurations requires additional constraints. We define the following constraints for exploiting the candidate solution's associations. {\color{blue}For each commit in the candidate solution, we perform exploitation in two directions. For the first direction, at least 80\,\% of  configuration options associated with a commit are required to be selected, whereas, for the second direction, at least 80\,\% configuration options not associated with a commit are required to be selected. This way, we limit variation among configuration options. The first direction aims at enabling additional configuration options to search for previously unseen associations (under-approximation). The second direction aims at the opposite: We limit variation to those configuration options already associated with a commit and therefore search for configuration options that are falsely associated (over-approximation). We use the limit of 80\,\% to allow a broader scope of valid configurations to be sampled by a constraint solver. For each new configuration, we assess performance for randomly selected commits.}
